Classical mechanics as the high-entropy limit \\ of quantum mechanics

### 2025-12-02 Articles/2024-ClassicalLimit/classicalLimitArticle.tex
ClassicalLimit: addressing more reviewer commetns

The key insight is that \textbf{high entropy is the one and only requirement}. We do not require a particular interpretation of quantum mechanics, of entropy or a particular mechanism responsible for the high entropy. Any will do.\footnote{The only requirement for entropy is that it matches the experimentally measured one and that its maximization corresponds to the correct experimental quantities. In this vein, the Shannon/Gibbs entropy of classical mechanics and the von Neumann entropy of quantum mechanics are simply the tools we use to make predictions within the respective theories. The Boltzmann entropy coincides with the Shannon/Gibbs entropy for a uniform distribution. }

@@

Mathematically, the trace distance $T(\rho,\sigma)$ between two mixed states $\rho$ and $\sigma$ is contractive under a completely positive trace-preserving (CPTP) map $M$. That is, $T(M(\rho),M(\sigma)) \leq T(\rho,\sigma)$~\cite{nielsen2010quantum}. In the finite-dimensional case, we can have a CPTP map that increases the entropy of all states except that of the maximally mixed state, which is a fixed point.\footnote{This is a particular instance of mixing/relaxing ergodic quantum channels where the fixed point is the maximally mixed state~\cite{Burgarth2012ErgodicAM}.}  Under such a map, all pure states are mapped to mixed states, orthogonal states are mapped to non-orthogonal states and the map is strictly contractive. Moreover, even if observable do not commute in general, they commute for the maximally mixed state $\rho = I/n$ in the following sense:
\begin{equation}
	\<AB\>_{\rho} = \tr(AB\rho) = \tr(ABI/n) = \tr(AB)/n = \tr(BA)/n = \tr(BAI/n)= \tr(BA\rho) = \<BA\>_{\rho}
\end{equation}
Observables for a state close enough to the maximally mixed state, then, will ``almost commute'' in the sense that the expectation of different permutations of their product will be small.

@@

What we described here is just a geometric property of the space of quantum mixed states, and it is independent of the source of entropy.

@@

The failure of classical statistical mechanics to predict the black-body radiation spectrum was one of the key drivers for the development of quantum theory. The spectrum predicted by classical mechanics is given by the Rayleigh–Jeans law,
\begin{equation}
    \frac{2 \nu^2 k_rext{B} T}{c^2} = \frac{2 \nu^2}{c^2 \beta},
\end{equation}
where $\beta = \frac{1}{k_\text{B} T}$. The one predicted by quantum mechanics is given by Planck's law,
\begin{equation}
    \frac{ 2 h \nu^3}{c^2} \frac 1{\exp\left(\frac{h\nu}{k_\mathrm B T}\right) - 1} = \frac{ 2 h \nu^3}{c^2} \frac 1{e^{h \beta \nu} - 1}.
\end{equation}
It is well known that the two agree for small values of $\nu$ since $e^x \simeq 1 + x + O(x^2)$ for $x\ll 1$.
\begin{equation}
    \frac{ 2 h \nu^3}{c^2} \frac 1{e^{h \beta \nu} - 1} = \frac{ 2 h \nu^3}{c^2} \frac 1{1 + h \beta \nu + O\left(\nu^2\right) - 1} \simeq \frac{2 \nu^2}{c^2 \beta}.
\end{equation}

@@

@@

It should be clear that this investigation does not exhaust all possible insights and links to the now vast literature on open quantum systems, quantum statistical mechanics, quantum information, quantum sensing and so on. For example, entropy and entropy increase play a role in several approaches: coarse graining(\href{https://journals.aps.org/prl/abstract/10.1103/PhysRevLett.99.180403}{CITE}), decoherence(CITE), stochastic processes (CITE \href{https://journals.aps.org/prl/abstract/10.1103/PhysRevLett.129.050401}{Quantum Fokker-Planck Master Equation for Continuous Feedback Control}) and information theoretical approaches to name a few. On a more technical level, since both the Husimi Q and the Wigner function converge to the classical distribution, the Wehrl entropy, which is the Shannon entropy calculated on the Husimi Q, will converge to the Gibbs entropy. One may ask, then, whether a high Wehrl entropy is already an indicator of classicality, and, if so, it would give another insight as to why the anti-normal ordering is important. Lastly, the role of entropic ideas is well established in historic literature. It would not be possible here to comment on all of them, and the more we mention, the more the omitted mentions are felt. Therefore we had to only explore the most important elements to establish the limit mathematically and under the smallest conceptual footprint, also as to not distract from the main point of the paper.

@@

We have seen that classical mechanics can be recovered as the high-entropy limit of quantum mechanics. That is, states of higher and higher entropy are better and better approximated by classical distributions over phase space. This approach to the classical limit is independent of mechanism and interpretation, as it does not matter how the entropy is increased or what one believes quantum states to represent: as long as the description is in terms of mixed states of sufficiently high entropy, classical mechanics applies. The approach fits naturally with experimental considerations and other approaches, such as decoherence, and recovers the established mathematical recipe, which is the group contraction for $\hbar \to 0$. Physically, this can be understood as taking the entropy of pure states to minus infinity, which is equivalent to saying that the relative entropy of mixed states goes to plus infinity. That is, in the same way that $c \to \infty$ should be understood as $v \ll c$, $\hbar \to 0$ should be understood as $S \gg 0$, as physically we cannot take limits of physical constants. This gives a reasonable and precise account of the classical limit: in the same way that non-relativistic mechanics applies to low speed, classical mechanics applies to high entropy.

>>>>

The key insight is that \textbf{high entropy is the one and only requirement}. We do not require a particular interpretation of quantum mechanics, of entropy or a particular mechanism responsible for the high entropy. Any will do. \footnote{The only requirement for entropy is that it matches the experimentally measured one and that its maximization corresponds to the correct experimental quantities. In this vein, the Shannon/Gibbs entropy of classical mechanics and the von Neumann entropy of quantum mechanics are simply the tools we use to make predictions within the respective theories. The Boltzmann entropy coincides with the Shannon/Gibbs entropy for a uniform distribution. }

@@

Mathematically, the trace distance $T(\rho,\sigma)$ between two mixed states $\rho$ and $\sigma$ is contractive under a completely positive trace-preserving (CPTP) map $M$. That is, $T(M(\rho),M(\sigma)) \leq T(\rho,\sigma)$~\cite{nielsen2010quantum}. In the finite-dimensional case, we can have a CPTP map that increases the entropy of all states except that of the maximally mixed state, which is a fixed point.\footnote{This is a particular instance of mixing/relaxing ergodic quantum channels where the fixed point is the maximally mixed state~\cite{Burgarth2012ErgodicAM}.}  Under such a map, all pure states are mapped to mixed states, orthogonal states are mapped to non-orthogonal states and the map is strictly contractive. Moreover, even if observables do not commute in general, they commute for the maximally mixed state $\rho = I/n$ in the following sense:
\begin{equation}
	\<AB\>_{\rho} = \tr(AB\rho) = \tr(ABI/n) = \tr(AB)/n = \tr(BA)/n = \tr(BAI/n)= \tr(BA\rho) = \<BA\>_{\rho}
\end{equation}
Observables for a state close enough to the maximally mixed state, then, will ``almost commute'' in the sense that the difference in the expectation of different permutations of their product will be small.

@@

@@

The failure of classical statistical mechanics to predict the black-body radiation spectrum was one of the key drivers for the development of quantum theory. The spectrum predicted by classical mechanics is given by the Rayleigh–Jeans law,
\begin{equation}
    \frac{2 \nu^2 k_\text{B} T}{c^2} = \frac{2 \nu^2}{c^2 \beta},
\end{equation}
where $\beta = \frac{1}{k_\text{B} T}$. The one predicted by quantum mechanics is given by Planck's law,
\begin{equation}
    \frac{ 2 h \nu^3}{c^2} \frac 1{\exp\left(\frac{h\nu}{k_\mathrm B T}\right) - 1} = \frac{ 2 h \nu^3}{c^2} \frac 1{e^{h \beta \nu} - 1}.
\end{equation}
It is well known that the two agree for small values of $\nu$ since $e^x \simeq 1 + x + O(x^2)$ for $x\ll 1$.
\begin{equation}
    \frac{ 2 h \nu^3}{c^2} \frac 1{e^{h \beta \nu} - 1} = \frac{ 2 h \nu^3}{c^2} \frac 1{1 + h \beta \nu + O\left(\nu^2\right) - 1} \simeq \frac{2 \nu^2}{c^2 \beta}.
\end{equation}

@@

While we concentrated on the one dimensional case, the result can be generalized to multiple dimensions by performing a similar stretching operation on all degrees of freedom simultaneously. That is, the entropy, as it increases, it spreads equally not just within each degree of freedom, but also across degrees of freedom. This would give us $[T_Q(X^i),T_Q(P_i)] = \lambda \imath \hbar$, so that the same stretching factor is associated to all the appearances of $\hbar$ in the commutation relationship.

@@

It should be clear that this investigation does not exhaust all possible insights and links to the now vast literature on open quantum systems, quantum statistical mechanics, quantum information, quantum sensing and so on. For example, entropy and entropy increase play a role in several approaches: coarse graining(\href{https://journals.aps.org/prl/abstract/10.1103/PhysRevLett.99.180403}{CITE}), decoherence(CITE 1 2 5 ), stochastic processes (CITE \href{https://journals.aps.org/prl/abstract/10.1103/PhysRevLett.129.050401}{Quantum Fokker-Planck Master Equation for Continuous Feedback Control}) and information-theoretic and resource-theoretic approaches (cite 3 4 6 7) to name a few. On a more technical level, since both the Husimi Q and the Wigner function converge to the classical distribution, the Wehrl entropy, which is the Shannon entropy calculated on the Husimi Q, will converge to the Gibbs entropy. One may ask, then, whether a high Wehrl entropy is already an indicator of classicality, and, if so, it would give another insight as to why the anti-normal ordering is important. Lastly, the role of entropic ideas is well established in historic literature. It would not be possible here to comment on all of them, and the more we mention, the more the omitted mentions are felt. Therefore we had to only explore the most important elements to establish the limit mathematically and under the smallest conceptual footprint, also as to not distract from the main point of the paper.

@@

We have seen that classical mechanics can be recovered as the high-entropy limit of quantum mechanics. That is, states of higher and higher entropy are better and better approximated by classical distributions over phase space and, if the entropy increase is done uniformly, unitary evolution can be approximated with classical Hamiltonian evolution. This approach to the classical limit is independent of mechanism and interpretation, as it does not matter how the entropy is increased or what one believes quantum states to represent: as long as the description is in terms of mixed states of sufficiently high entropy, classical mechanics applies. The approach fits naturally with experimental considerations and other approaches, such as decoherence, leading to an interpretation that is  thermodynamically meaningful, representation-independent, and stable under coarse-graining. It also recovers the established mathematical recipe, which is the group contraction for $\hbar \to 0$.\footnote{As pointed out by a reviewer, the $\hbar \to 0$ limit breaks down for chaotic or non-analytic systems, while the high entropy limit will always exist. Therefore there may be an additional concrete advantage to our approach over the standard one.} Physically, this can be understood as taking the entropy of pure states to minus infinity, which is equivalent to saying that the relative entropy of mixed states goes to plus infinity. That is, in the same way that $c \to \infty$ should be understood as $v \ll c$, $\hbar \to 0$ should be understood as $S \gg 0$, as physically we cannot take limits of physical constants. This gives a reasonable and precise account of the classical limit: in the same way that non-relativistic mechanics applies to low speed, classical mechanics applies to high entropy.
